---
title: "combining study findings"
format: html
---

# combinding study findings

this scripts will take the p-values from differential expression and differential isoform expression

for joining p-values of the same tissue and expression we will use edgingtons method. it will be performed by:

`metap::sump(c(result1,results2)$p)`

For studies of different tissues or methods, we will use Fishers method, which prioritizes the smallest p-value, as we in the end want genes that are significant for any of the tested properties (tissue, expression, isoform ...)

`p_join_fisher = metap::sumlog(c(pvalue_deseq2,pvalue_dexseq))$p )`

## load DEseq2 results and DEXseq results

```{r}
load("/home/projects/exam_2025_22140_22150/group_2/data/GSE67333_DXD.Rdata")
load("/home/projects/exam_2025_22140_22150/group_2/data/GSE173955_result_objects.Rdata")

```

## combinding p-values for same studies

### combinding DE

for studies performing the same test on the same tissue, we will combine the p-values using the edgingtons method.

```{r}

# full_join datasets

# perform p-value adjustment

```
